---
title: "BIOS 663 Group Project - Wage Data"
author: "Wage Groupies"
date: "5/3/2021"
output: html_document
---

```{r,message=FALSE,warning=FALSE}
#applicable packages
library(tidyverse)
library(dplyr)
library(mosaic)
library(car)
#load data - Andrew
wages <- read.csv("~/Bios-663-Project/CPS_85_Wages.txt", sep="")
#load data - Others
```

### Presentation (15 minutes)
1. description of the data
2. description and justification of the analysis 
3. analysis results (including relevant descriptive statistics, tables, and graphs)
4. conclusions and discussions

### EDA

```{r}
#head and missing values
head(wages)
#no missing values in the dataset
a <- sum(complete.cases(wages))
b <- nrow(wages)
print(a)
print(b)
```

```{r,include=FALSE}
##counts of binary variables
#South
tally(wages$south)
#Sex
tally(wages$sex)
#Union
tally(wages$union)
#Race
tally(wages$race)
#Occupation
tally(wages$occupation)
#Sector
tally(wages$sector)
#Marriage
tally(wages$marriage)
```

South|Count
----|----
Lives in South|156
Lives Elsewhere|378

Sex|Count
----|----
Male|289
Female|245

Union|Count
----|----
Union Member|96
Not Union Member|438

Race|Count
----|----
White|440
Hispanic|27
Other|67

Occupation|Count
----|----
Management|55
Sales|38
Clerical|97
Service|83
Professional|105
Other|156

Sector|Count
----|----
Manufacturing|99
Construction|24
Other|411

Marital Status|Count
----|----
Married|350
Unmarried|184


```{r}
##eda histograms of all numeric covariates
#education
wages %>% ggplot(aes(x=education)) + 
  geom_histogram(binwidth=1) + 
  labs(x='Years of Education', y='Count', title = 'Count of Years of Education')
#experience
wages %>% ggplot(aes(x=experience)) + 
  geom_histogram(binwidth=5) + 
  labs(x='Years of Experience', y='Count', title = 'Count of Years of Experience')
#age
wages %>% ggplot(aes(x=age)) + 
  geom_histogram(binwidth=5) + 
  labs(x='Years of Age', y='Count', title = 'Count of Years of Education')
#wage
wages %>% ggplot(aes(x=wage)) + 
  geom_density(fill="dodgerblue1",alpha=0.5) + 
  labs(x='Wage (dollars/hour)', y='Count', title = 'Count of Wages (dollars/hour)')
```
```{r}
mean(wages$wage)
median(wages$wage)
range(wages$wage)
favstats(wages$wage)
```



### Interesting Metrics (group)

```{r}
#proportion of wages by sex
a <- wages %>% filter(sex == 1) %>% sum(wages$wage)
b <- wages %>% filter(sex == 0) %>% sum(wages$wage)
c <- a+b
prop_men <- a/c
prop_women <- b/c
print(prop_men)
print(prop_women)
#calculate proportion of men and women in data
d <- count(wages$sex==1)/nrow(wages)
e <- count(wages$sex==0)/nrow(wages)
print(d)
print(e)
```

- Women make up 47% of wages, Men make up 53% of wages but Women make up 46% of observations and men make up 54% of observations.

```{r}
#average wage by sex
tapply(wages$wage, wages$sex, mean)
tapply(wages$wage, wages$sex, median)

ggplot(wages)+geom_boxplot(aes(y=wage,x=as.factor(sex),color=as.factor(sex))) + labs(title="Hourly Wage by Sex",x="Sex (Male(0),Female(1))",y="Wage ($/hour)",color="Sex") 
```

```{r}
#average wage by education level
tapply(wages$wage, wages$education, mean)
tapply(wages$wage, wages$education, median)

```

```{r}
#average wage by residence area
tapply(wages$wage, wages$south, mean)
tapply(wages$wage, wages$south, median)

ggplot(wages)+geom_boxplot(aes(y=wage,x=as.factor(south),color=as.factor(south)))
```


```{r}
#average wage by experience level
tapply(wages$wage, wages$experience, mean)
tapply(wages$wage, wages$experience, median)
```

```{r}
#average wage by union classification
tapply(wages$wage, wages$union, mean)
tapply(wages$wage, wages$union, median)
```

```{r}
#average wage by age
tapply(wages$wage, wages$age, mean)
tapply(wages$wage, wages$age, median)
```

```{r}
#average wage by race level
tapply(wages$wage, wages$race, mean)
tapply(wages$wage, wages$race, median)
```

```{r}
#average wage by occupation
tapply(wages$wage, wages$occupation, mean)
tapply(wages$wage, wages$occupation, median)
```

```{r}
#average wage by sector
tapply(wages$wage, wages$sector, mean)
tapply(wages$wage, wages$sector, median)
```

```{r}
#average wage by marital status
tapply(wages$wage, wages$marriage, mean)
tapply(wages$wage, wages$marriage, median)
```

### Backward Selection Regression Model

```{r}
#full model
model.1 <- lm(wage ~ . ,data = wages)
summary(model.1)
```

```{r}
vif(model.1)
```

```{r}
#step-wise backward model selection beginning with max model
step(model.1, direction='backward')
```

```{r}
#reduced model
model.2 <- lm(wage ~ education+south+sex+experience+union+race+sector,data = wages)
summary(model.2)
```


```{r}
# Numerical predictors:
# Age, experience, education
colnames(wages)

wages %>% 
  ggplot(.) +
  geom_point(aes(y = wage, x = age, color = factor(race))) +
  facet_grid(union~sex)
```

### ANOVA test for wage by sex

```{r}

```

